Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion

### 2023-08-02 Manuscript.Rmd
edits: prosimodistal bias

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments (Fig. 2). When rating the cold thermode (Exp. 1, Fig. 2A), participants reported a significantly reduced subjective experience of cold for TGI, but not non-TGI, stimuli applied within a dermatome compared to across dermatomes (stimulation by dermatome interaction: $\beta$ = `r stats_model_cold_exp1$beta[6]`, `r make_pvalue(stats_model_cold_exp1$p[6])`). Further, participants rated both TGI and non-TGI stimuli as warmer within vs across dermatome (stimulation by dermatome interaction: $\beta$ = `r stats_model_warm_exp1$beta[6]`, `r make_pvalue(stats_model_warm_exp1$p[6])`; dermatome main effect: $\beta$ = `r stats_model_warm_exp1$beta[2]`, `r make_pvalue(stats_model_warm_exp1$p[2])`). Finally, participants reported no significant modulation of burning ratings depending on the dermatome condition (stimulation by dermatome interaction: $\beta$ = `r stats_model_burn_exp1$beta[6]`, `r make_pvalue(stats_model_burn_exp1$p[6])`; dermatome main effect: $\beta$ = `r stats_model_burn_exp1$beta[2]`, `r make_pvalue(stats_model_burn_exp1$p[2])`). Overall, these results indicated that when participants assessed the cold thermode, the greatest modulation in TGI perception was related to cold perception, with within-dermatome TGI stimuli perceived as the least cold.

@@

The findings suggest that the thermosensory quality of the reference thermode (cold or warm) showed differential sensitivity to thermosensory and painful aspects of the TGI experience, and collectively suggested that both the thermosensory enhancement and illusory pain components of the illusion are modulated at the spinal cord level.
This interpretation is consistent with a previous study using a similar dermatome manipulation, but a distinct method to quantify TGI perception [@fardo_organization_2018], as well as another study showing modulation of heat and pain ratings of TGI stimuli by conditioned pain modulation [@harper_conditioned_2017].
All together these results support the role of spinal processes in the generation of distinct perceptual aspects of the TGI, and highlight the importance of the reference stimulus when assessing TGI perception.

## Undetected proximodistal bias in TGI

Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli presented on the forearm [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007].
In our study, we sought to investigate whether this effect of distal inhibition would also influence TGI perception within a single dermatome (Fig. 3).
We tested this effect with a two way interaction between condition (control / TGI) and the location of the cold thermode when it was placed within dermatome (distal / proximal).
In the first experiment with the cold thermode as the reference we found no statistical significant interaction effect for the cold ratings ($\beta$ = `r stats_model_cold_exp1$beta[8]`, `r make_pvalue(stats_model_cold_exp1$p[8])`, however with a strong main effect of dermatome ($\beta$ = `r stats_model_cold_exp1$beta[4]`, `r make_pvalue(stats_model_cold_exp1$p[4])`) For the warm ratings we found no interaction ($\beta$ = `r stats_model_warm_exp1$beta[8]`, `r make_pvalue(stats_model_warm_exp1$p[8])`), nor main effect of dermatome ($\beta$ = `r stats_model_warm_exp1$beta[4]`, `r make_pvalue(stats_model_warm_exp1$p[4])`) For the burning ratings We found no significant interaction ($\beta$ = `r stats_model_burn_exp1$beta[8]`, `r make_pvalue(stats_model_burn_exp1$p[8])`), but a significant main effect ( $\beta$ = `r stats_model_burn_exp1$beta[4]`, `r make_pvalue(stats_model_burn_exp1$p[4])`)

For the second experiment with the warm thermode as the reference we found no interaction no interaction ($\beta$ = `r stats_model_burn_exp2$beta[8]`, `r make_pvalue(stats_model_burn_exp2$p[8])`), nor main effect ($\beta$ = `r stats_model_burn_exp2$beta[4]`, `r make_pvalue(stats_model_burn_exp2$p[4])`) for the burning ratings.
For the warm ratings we found no interaction ($\beta$ = `r stats_model_warm_exp2$beta[8]`, `r make_pvalue(stats_model_warm_exp2$p[8])`), , nor main effect of dermatome ($\beta$ = `r stats_model_warm_exp2$beta[4]`, `r make_pvalue(stats_model_warm_exp2$p[4])`) and for the cold ratings no interaction ($\beta$ = `r stats_model_cold_exp2$beta[8]`, `r make_pvalue(stats_model_cold_exp2$p[8])`) but a main effect ($\beta$ = `r stats_model_cold_exp2$beta[4]`, `r make_pvalue(stats_model_cold_exp2$p[4])`).

@@

>>>>

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments.
When rating the cold thermode (Exp. 1, Fig. 2A), participants reported a reduced subjective experience of cold for TGI, but not non-TGI, stimuli applied within a dermatome compared to across dermatomes (stimulation by dermatome interaction: $\beta$ = `r stats_model_cold_exp1$beta[6]`, `r make_pvalue(stats_model_cold_exp1$p[6])`), alongside an increased subjective experience of warmth for both TGI and non-TGI stimuli (stimulation by dermatome interaction: $\beta$ = `r stats_model_warm_exp1$beta[6]`, `r make_pvalue(stats_model_warm_exp1$p[6])`; dermatome main effect: $\beta$ = `r stats_model_warm_exp1$beta[2]`, `r make_pvalue(stats_model_warm_exp1$p[2])`).
Further, participants reported no significant modulation of burning ratings depending on the dermatome condition (stimulation by dermatome interaction: $\beta$ = `r stats_model_burn_exp1$beta[6]`, `r make_pvalue(stats_model_burn_exp1$p[6])`; dermatome main effect: $\beta$ = `r stats_model_burn_exp1$beta[2]`, `r make_pvalue(stats_model_burn_exp1$p[2])`).
Overall, these results indicated that when participants assessed the cold thermode, the greatest modulation in TGI perception was related to cold perception, with within-dermatome TGI stimuli perceived as the least cold.
While the modulation of cold perception was specific for TGI, increased warmth was reported irrespective of whether the cold thermode was paired with a warm (TGI stimuli) or neutral temperature (non-TGI stimuli) within a dermatome.
These findings are in line with the notion that TGI can be considered a misperception of cold [@fardo_organization_2018; @fardo_beyond_2020].

@@

The findings suggest that the thermosensory quality of the reference thermode (cold or warm) showed differential sensitivity to thermosensory and painful aspects of the TGI experience, and collectively suggested that both qualitative components of the illusion are modulated at the spinal cord level.
This interpretation is consistent with a previous study using a similar dermatome manipulation, but a distinct method to quantify TGI perception [@fardo_organization_2018], as well as another study showing modulation of heat and pain ratings of TGI stimuli by conditioned pain modulation [@harper_conditioned_2017].
All together these results support the role of spinal processes in the generation of distinct perceptual aspects of the TGI, and highlight the importance of the reference stimulus when assessing thermosensory and burning components of TGI perception.

## Proximodistal bias in cold perception

Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007].
We tested the generazability of this distal inhibition effect to the perception of mild temperatures in TGI and non-TGI stimuli within single dermatomes (Fig. 3).
We found that cold perception was modulated by the proximodistal location of the reference thermode, with enhanced cold ratings when the reference thermode was located more proximally, irrespective of whether the stimulus was TGI or non-TGI, in both Exp. 1 (stimulation by dermatome interaction: $\beta$ = `r stats_model_cold_exp1$beta[8]`, `r make_pvalue(stats_model_cold_exp1$p[8])`; dermatome main effect: $\beta$ = `r stats_model_cold_exp1$beta[4]`, `r make_pvalue(stats_model_cold_exp1$p[4])`) and Exp. 2 (stimulation by dermatome interaction: $\beta$ = `r stats_model_cold_exp2$beta[8]`, `r make_pvalue(stats_model_cold_exp2$p[8])`; dermatome main effect: $\beta$ = `r stats_model_cold_exp2$beta[4]`, `r make_pvalue(stats_model_cold_exp2$p[4])`).
These findings extend the notion of distal inhibition to innocuous cold perception.

@@

Cuts

From "Proximodistal bias in cold perception":

Experiment 1: warm ratings (stimulation by dermatome interaction: $\beta$ = `r stats_model_warm_exp1$beta[8]`, `r make_pvalue(stats_model_warm_exp1$p[8])`; dermatome main effect: $\beta$ = `r stats_model_warm_exp1$beta[4]`, `r make_pvalue(stats_model_warm_exp1$p[4])`) or burning ratings (stimulation by dermatome interaction: $\beta$ = `r stats_model_burn_exp1$beta[8]`, `r make_pvalue(stats_model_burn_exp1$p[8])`; dermatome main effect: $\beta$ = `r stats_model_burn_exp1$beta[4]`, `r make_pvalue(stats_model_burn_exp1$p[4])`)

Experiment 2: warm ratings (stimulation by dermatome interaction: $\beta$ = `r stats_model_warm_exp2$beta[8]`, `r make_pvalue(stats_model_warm_exp2$p[8])`); dermatome main effect: $\beta$ = `r stats_model_warm_exp2$beta[4]`, `r make_pvalue(stats_model_warm_exp2$p[4])`) or burning ratings (stimulation by dermatome interaction: $\beta$ = `r stats_model_burn_exp2$beta[8]`, `r make_pvalue(stats_model_burn_exp2$p[8])`); dermatome main effect: $\beta$ = `r stats_model_burn_exp2$beta[4]`, `r make_pvalue(stats_model_burn_exp2$p[4])`).
